Anomaly detection with kernel density estimation on manifolds

### 2022-03-29 paper/paper.Rmd
Corrections

In topology, a *homeomorphism* is a bijective map between two topological spaces that is continuous in both directions. 
A *Hausdorff space* is a topological space where any two distinct points can be separated by disjoint neighborhoods.
A $d$-dimensional (topological) *manifold* $M$ is a connected Hausdorff space $(M, \mathcal{T}_M)$ where the neighborhood $U$ for each point $p$ is homeomorphic to an open subset $V$ of the Euclidean space $\mathbb{R}^d$. Such a homeomorphism $\varphi: U \rightarrow V$ and $U$ gives a (coordinate) *chart*, denoted as $(U, \varphi)$, with the corresponding local coordinates $(x_1(p),\dots, x_d(p)) := \varphi(p)$. A collection of charts $\{U_\alpha, \varphi_\alpha\}$ ranging over the manifold $M$ is called an *atlas*, denoted as $\mathcal{A}$.
The manifold $M$ is a *differentiable manifold* if there exists an atlas of $M$, $\{U_\alpha, \varphi_\alpha\}$, such that the *transition maps* between any two charts,
$$
\varphi_\beta \circ \varphi_\alpha^{-1}: \varphi_\alpha(U_\alpha \cap U_\beta) \rightarrow \varphi_\beta(U_\alpha \cap U_\beta),
$$
are differentiable of class $C^\infty$ (smooth).

Let $\varphi$ be an injective map $E \rightarrow \varphi(E)$. Then $\varphi$ is an embedding of $E$ into $M$ if and only if $\varphi: E \rightarrow \varphi(E)$ is a homeomorphism. $\varphi(N)$ is called an embedded submanifold of $M$ with the subspace topology.

@@

For any $p \in M$, let $\gamma_1: (-\epsilon_1, \epsilon_1)\rightarrow M$ and $\gamma_2: (-\epsilon_2, \epsilon_2)\rightarrow M$ be two smooth curve through $p$, i.e. $\gamma_1(0) = \gamma_2(0) = p$. $\gamma_1$ and $\gamma_2$ are *equivalent* if and only if there exists a chart, $(U,\varphi)$, at $p$ such that
$$
(\varphi \circ \gamma_1)^\prime(0) = (\varphi \circ \gamma_2)^\prime(0).
$$
A *tangent vector* to $M$ at $p$ is any equivalent class of the differentiable curves initialized at $p$. 
The set of all tangent vectors at $p$ defines the *tangent space* of $M$ at $p$, denoted as $T_pM$. The tangent space is vector space of dimension $d$, equal to the dimension of $M$, and it does not depend on the charts $\varphi$ locally at $p$. The collection of all tangent spaces defines the *tangent bundle* $TM = \cup_{p \in M}T_pM$.
Tangent vectors can also be seen as the directional derivatives at $p$. For a given coordinate chart $\varphi=(x_1,\dots,x_d)$, the tangent vectors defining partial derivatives are denoted as $\frac{\partial}{\partial x_1}(p),\dots,\frac{\partial}{\partial x_d}(p)$, which define a *basis* of the tangent space $T_pM$.

@@

@@

<!-- embedding pushforward metric [@Perrault-Joncas2013-pq] -->

The Riemannian metric $g$ can be used to define the norm of a vector $u$, $\|u\| = \sqrt{<u,v>_g}$ and the angle between two vectors $u$ and $v$, $\cos\theta = \frac{<u,v>_g}{\|u\| \|v\|}$, which are the geometric quantities induced by $g$.
$g$ could also be used to define the line element $dl^2 = g_{ij}dx_idx_j$ and the volume element $dV_g = \sqrt{\det(g)}dx_1\dots dx_d$, where $(x_1,\dots,x_d)$ are the local coordinates of the chart $(U, \varphi)$. 
For a curve $\gamma: I \rightarrow M$, the length of the curve is 
$$l(\gamma) = \sqrt{\int_0^1 \|\gamma\prime(t)\|^2_g dt} = \sqrt{\int_0^1 g_{ij} \frac{dx_i}{dt} \frac{dx_j}{dt} dt}$$,
where $\gamma(I) \subset U$. The volume $W \subset U$ is defined as
$$
Vol(W) = \int_W \sqrt{\det(g)}dx_1 \dots dx_d,
$$
which is also called the *Riemannian measure* on $M$.

The *geodesics* of $M$ are the smooth curves that locally joins the points along the shortest path on the manifold. A curve $\gamma: I \rightarrow M$ is a geodesic if for all indices $i,j,k$,
the second-order ordinary differential equation is satisfied,
$$
\frac{d^2 x^i}{dt^2} + \Gamma^i_{jk} \frac{d x_j}{dt} \frac{dx_k}{dt} = 0,
$$
where $\Gamma^i_{jk}$ is the *Christoffel symbol* defined by
$$
\Gamma^i_{jk} = \frac{1}{2} \sum_l g_{il} (\frac{\partial g_{il}}{\partial x_k} 
+ \frac{\partial g_{kl}}{\partial x_j} - \frac{\partial g_{jk}}{\partial x_l}).
$$
The geodesics have a constant speed with norm $\| \gamma^\prime(t) \|$, and they are the local minimizers of the arc length functional $l:\gamma \rightarrow \int_0^1 \| \gamma^\prime(t) \| dt$ when the curves are defined over the interval $[0,1]$. The *geodesic distance* $d_g$ is the length of the shortest geodesic between two points on the manifold. 
For a point $p \in M$, when the geodesic distance starting at $p$ is not minimized, we call such set of points the *cut locus* of $p$, and the *injectivity radius* at $p \in M$ is the distance to the cut locus. The injectivity radius of $(M,g)$, $\textit{inj}_gM$ is the infimum of the injectivity radii over all points on the manifold.

## Exponential map and logarithmic map

>>>>

In topology, a *homeomorphism* is a bijective map between two topological spaces that is continuous in both directions. 
A *Hausdorff space* is a topological space where any two distinct points can be separated by disjoint neighborhoods.
A $d$-dimensional (topological) *manifold* $M$ is a connected Hausdorff space $(M, \mathcal{T}_M)$ where the neighborhood $U$ for each point $p$ is homeomorphic to an open subset $V$ of the Euclidean space $\mathbb{R}^d$. Such a homeomorphism $\varphi: U \rightarrow V$ together with $U$ gives a (coordinate) *chart*, denoted as $(U, \varphi)$, with the corresponding local coordinates $(x_1(p),\dots, x_d(p)) := \varphi(p)$. A collection of charts $\{U_\alpha, \varphi_\alpha\}$ ranging over the manifold $M$ is called an *atlas*, denoted as $\mathcal{A}$.
The manifold $M$ is a *differentiable manifold* if there exists an atlas of $M$, $\{U_\alpha, \varphi_\alpha\}$, such that the *transition maps* between any two charts,
$$
\varphi_\beta \circ \varphi_\alpha^{-1}: \varphi_\alpha(U_\alpha \cap U_\beta) \rightarrow \varphi_\beta(U_\alpha \cap U_\beta),
$$
are differentiable of class $C^\infty$ (smooth).

Let $\varphi$ be an injective map $E \rightarrow \varphi(E)$. Then $\varphi$ is an embedding of $E$ into $M$ if and only if $\varphi: E \rightarrow \varphi(E)$ is a homeomorphism, and $\varphi(E)$ is called an embedded submanifold of $M$ with the subspace topology.

@@

For any $p \in M$, let $\gamma_1: (-\epsilon_1, \epsilon_1)\rightarrow M$ and $\gamma_2: (-\epsilon_2, \epsilon_2)\rightarrow M$ be two smooth curve through $p$, i.e. $\gamma_1(0) = \gamma_2(0) = p$. $\gamma_1$ and $\gamma_2$ are *equivalent* if and only if there exists a chart, $(U,\varphi)$, at $p$ such that
$$
(\varphi \circ \gamma_1)^\prime(0) = (\varphi \circ \gamma_2)^\prime(0).
$$
A *tangent vector* to a manifold $M$ at point $p$, $v_p$, is any equivalent class of the differentiable curves initialized at $p$. 
The set of all tangent vectors at $p$ defines the *tangent space* of $M$ at $p$, denoted as $T_pM$. The tangent space is a vector space of dimension $d$, equal to the dimension of $M$, and it does not depend on the chart $\varphi$ locally at $p$. The collection of all tangent spaces defines the *tangent bundle* $TM = \cup_{p \in M}T_pM$.
Tangent vectors can also be seen as the directional derivatives at $p$. For a given coordinate chart $\varphi=(x_1,\dots,x_d)$, the tangent vectors defining partial derivatives are denoted as $\frac{\partial}{\partial x_1}(p),\dots,\frac{\partial}{\partial x_d}(p)$, which define a *basis* of the tangent space.

@@

The Riemannian metric $g$ can be used to define the norm of a vector $u$, $\|u\| = \sqrt{<u,v>_g}$ and the angle between two vectors $u$ and $v$, $\cos\theta = \frac{<u,v>_g}{\|u\| \|v\|}$, which are the geometric quantities induced by $g$.
$g$ could also be used to define the line element $dl^2 = g_{ij}dx_idx_j$ and the volume element $dV_g = \sqrt{\det(g)}dx_1\dots dx_d$, where $(x_1,\dots,x_d)$ are the local coordinates of the chart $(U, \varphi)$. 
For a curve $\gamma: I \rightarrow M$, the length of the curve is 
$$l(\gamma) = \sqrt{\int_0^1 \|\gamma\prime(t)\|^2_g dt} = \sqrt{\int_0^1 g_{ij} \frac{dx_i}{dt} \frac{dx_j}{dt} dt}$$,
where $\gamma(I) \subset U$. The volume of $W \subset U$ is defined as
$$
Vol(W) = \int_W \sqrt{\det(g)}dx_1 \dots dx_d,
$$
which is also called the *Riemannian measure* on $M$.

The *geodesics* of $M$ are the smooth curves that locally joins the points along the shortest path on the manifold. A curve $\gamma: I \rightarrow M$ is a geodesic if for all indices $i,j,k$,
the second-order ordinary differential equation is satisfied,
$$
\frac{d^2 x_i}{dt^2} + \Gamma^i_{jk} \frac{d x_j}{dt} \frac{dx_k}{dt} = 0,
$$
where $\Gamma^i_{jk}$ is the *Christoffel symbol* defined by
$$
\Gamma^i_{jk} = \frac{1}{2} \sum_l g_{il} (\frac{\partial g_{il}}{\partial x_k} 
+ \frac{\partial g_{kl}}{\partial x_j} - \frac{\partial g_{jk}}{\partial x_l}).
$$
The geodesics have a constant speed with norm $\| \gamma^\prime(t) \|$, and they are the local minimizers of the arc length functional $l:\gamma \rightarrow \sqrt{\int_0^1 \| \gamma^\prime(t) \|_g^2 dt}$ when the curves are defined over the interval $[0,1]$. The *geodesic distance* $d_g$ is the length of the shortest geodesic between two points on the manifold. 
For a point $p \in M$, when the geodesic distance starting at $p$ is not minimized, we call such set of points the *cut locus* of $p$, and the *injectivity radius* at $p \in M$ is the distance to the cut locus. The injectivity radius of $(M,g)$, $\textit{inj}_gM$ is the infimum of the injectivity radii over all points on the manifold.

@@

<!-- embedding pushforward metric [@Perrault-Joncas2013-pq] -->                                                                                            
## Exponential map and logarithmic map
